---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rhinotypeR

<!-- badges: start -->
<!-- badges: end -->

The goal of rhinotypeR is to perform genotyping on rhinovirus sequences. 

- The package does the following:
  * Assigns genotypes to query sequences
  * Computes for pairwise distance among query sequences
  * Calculates pairwise distance between query and prototype sequences
  * Calculates overall genetic distance of query sequences
  
## Data Format
The input data is a reference fasta file. 

## Installation

You can install the development version of rhinotypeR from [GitHub](https://github.com) with:

``` r
devtools::install_github("omicscodeathon/rhinotyper/rhinotypeR")
```

## Example

How to run the functions using our test data:

```{r }
library(Rhinotyper)
```

```{r }
assignTypes(pathToRef, pathToQuery, model = "p-distance", threshold = 0.105)
```

```{r }
#Create an output directory in your current directory before running this function
getPrototypeSeqs(destinationFolder = "./output")
```

```{r }
allPrototypeDistances(RVBPrototype, pathToQuery, "p-distance")
```

```{r }
overallMeanDistance(RVAPrototype)
```
